---
title: "M5RR - Week 2 Assignment"
author: "Tnargz"
date: "5/27/2020"
output:
  html_document: default
---

1. Loading and preprocessing the data.  Show any code that is needed to load the data (i.e. \color{red}{\verb|read.csv()|}read.csv()) and process/transform the data (if necessary) into a format suitable for your analysis.
```{r load_packages, warning = FALSE, message = FALSE}
library(lubridate)
library(RCurl)
library(tidyverse)
library(ggplot2)
library(dplyr)
library(magrittr)
library(readr)
library(gridExtra)
library(knitr)
```

````{r preprocessing, echo = TRUE, warning = FALSE, message = FALSE}
FileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
ZipFile <- "./M5RR-Week2-Assignment/Data.zip"
FileDir <- "./M5RR-Week2-Assignment"
Path <- "./M5RR-Week2-Assignment/Data"
if (!file.exists(FileDir)){
  dir.create(FileDir)
}
download.file(FileUrl, file.path(ZipFile))
unzip(ZipFile, exdir = Path)
Data <- file.path(Path, "activity.csv")
activity <- read.csv(Data)
head(activity, 5)
```

2. total number of steps taken per day
```{r steps_per_day, echo = TRUE, warning = FALSE, message = FALSE}
TotalStepsPerDay <- activity %>%
  group_by(date) %>%
  summarise(stepsperday = sum(steps))
head(TotalStepsPerDay, 5)
```

2.1. What is mean total number of steps taken per day?
```{r mean_steps_per_day, echo = TRUE, warning = FALSE, message = FALSE}
TotalStepsPerDay_Mean <- TotalStepsPerDay %>%
  summarise(mean = mean(stepsperday, na.rm = TRUE))
TotalStepsPerDay_Mean
```

2.2. Make a histogram of the total number of steps taken each day.
```{r histogram_steps_per_day, echo = TRUE, warning = FALSE, message = FALSE}
ggplot(data = TotalStepsPerDay, aes(x = stepsperday), rm.na = TRUE) +
  geom_histogram(binwidth = 500, col = "black", fill = "red") +
  labs(x = "Total Daily Steps", y = "Frequency (binwidth 500)",
       title = "Total Number of Steps Taken Each Day")
```

2.3. Calculate and report the mean and median of the total number of steps taken per day
```{r mean_median_steps_per_day, echo = TRUE, warning = FALSE, message = FALSE}
TotalStepsPerDay_MeanMedian <- TotalStepsPerDay %>%
  summarise(mean = mean(stepsperday, na.rm = TRUE),
            median = median(stepsperday, na.rm = TRUE))
TotalStepsPerDay_MeanMedian
```

3. What is the average daily activity pattern?
```{r activity_pattern, echo = TRUE, warning = FALSE, message = FALSE}
activitypattern <- activity %>%
  group_by(interval) %>%
  summarise(mean_steps = mean(steps, na.rm = TRUE))
summary(activitypattern)
```

3.1. Make a time series plot (i.e. \color{red}{\verb|type = "l"|}type = "l") of the 5-minute interval (x-axis) and the average number of steps taken,,averaged across all days (y-axis).
```{r activity_pattern_plot, warning = FALSE, message = FALSE}
ggplot(data = activitypattern, aes(x = interval, y = mean_steps), rm.na = TRUE) +
  geom_line() +
  labs(x = "Interval (5 Minutes)", y = "Mean Steps",
       title = "Daily Activity Pattern")
```

3.2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r max_steps, echo = TRUE, warning = FALSE, message = FALSE}
MaxSteps <- activitypattern[order(-activitypattern$mean_steps),]
MaxSteps
```

4. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with \color{red}{\verb|NA|}NAs).
```{r missing_values, echo = TRUE, warning = FALSE, message = FALSE}
summary(activity)
```

4.1. Devise a strategy for filling in all of the missing values in the dataset.  The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.
```{r na_strategy, echo = TRUE, warning = FALSE, message = FALSE}
plot1 <- ggplot(activity, aes(activity$interval, activity$steps), rm.na = TRUE) +
  geom_boxplot()
plot2 <- ggplot(activity, aes(activity$steps), rm.na = TRUE) +
  geom_histogram()
grid.arrange(plot1, plot2, nrow = 2)
summary(activity)
```
  The median will be used due to the right skew nature of the data set.

4.2. Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r new_data, echo = TRUE, warning = FALSE, message = FALSE}
Na_to_Median <- function(x) replace(x, is.na(x), median(x, na.rm = TRUE))
activity2 <- activity %>% group_by(interval) %>% mutate(steps = Na_to_Median(steps))
summary(activity)
summary(activity2)
```

4.3. Make a histogram of the total number of steps taken each day and Calculateand report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?
```{r activity2_histogram, echo = TRUE, warning = FALSE, message = FALSE}
ggplot(data = activity2, aes(x = activity2$steps)) +
  geom_histogram(binwidth = 30, col = "black", fill = "yellow") +
  labs(x = "Total Daily Steps", y = "Frequency (binwidth 30)",
       title = "Total Number of Steps Taken Each Day")
summary(activity)
summary(activity2)
```
 Replacing the NA values with the median of the "activity" data set created a more narrow, right skewed data set.
 
5. Are there differences in activity patterns between weekdays and weekends? For this part the \color{red}{\verb|weekdays()|}weekdays() function may be of some help here. Use the dataset with the filled-in missing values for this part.

5.1 Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.

```{r weekday_weekend_variable, echo = TRUE, warning = FALSE, message = FALSE}
day_of_the_week <- wday(activity2$date)
f <- function(i) {
  day_to_category <- c('weekend',
                       'weekday',
                       'weekday',
                       'weekday',
                       'weekday',
                       'weekday',
                       'weekend')
  return (day_to_category[i])
}

activity2$time_of_week <- sapply(day_of_the_week, f)
str(activity2)
head(activity2, 5)

stepsperday_weekday <- activity2 %>%
  filter(time_of_week == 'weekday') %>%
  group_by(interval) %>%
  summarise(steps = mean(steps, na.rm = TRUE))
summary(stepsperday_weekday)

stepsperday_weekend <- activity2 %>%
  filter(time_of_week == 'weekend') %>%
  group_by(interval) %>%
  summarise(steps = mean(steps, na.rm = TRUE))
summary(stepsperday_weekend)
```

5.2. Make a panel plot containing a time series plot (i.e. \color{red}{\verb|type = "l"|}type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). See the README file in the GitHub repository to see an example of what this plot should look like using simulated data. Add a categorical variable for weekday vs weekend analysis
```{r weekday_weekend_plot, echo = TRUE, warning = FALSE, message = FALSE}
plot3 <- ggplot(data = stepsperday_weekday, aes(x = steps)) +
  geom_histogram(binwidth = 10, col = "black", fill = "white") +
  labs(x = "Total Daily Steps", y = "Frequency (binwidth 10)",
       title = "Total Steps Taken on Weekday")
plot4 <- ggplot(data = stepsperday_weekend, aes(x = steps)) +
  geom_histogram(binwidth = 10, col = "black", fill = "white") +
  labs(x = "Total Daily Steps", y = "Frequency (binwidth 10)",
       title = "Total Steps Taken on Weekend")
grid.arrange(plot3, plot4, nrow = 2)
```